PanGraph: scalable bacterial pan-genome graph construction

### 2022-09-29 main.tex
Update on Overleaf.

We quantified the accuracy of PanGraph algorithm by computing the displacement of the inferred breakpoints relative to their known locus stored from the evolutionary simulation (cf. SI Section IIB for details). We evaluate this displacement by generating datasets with different rates of mutation $\mu$ and HGT $h$. For each $(h,\mu)$ pair we perform 25 different simulations, and build pangenome graphs using three different options for the alignment kernel, namely \textit{minimap2} \cite{li2018minimap2} with \textit{asm10} and \textit{asm20} option and \textit{mmseqs2} \cite{steinegger2017mmseqs2}.
Critically, we found that the accuracy was \emph{independent} of the rate of horizontal gene transfer $h$ and thus underlying graph complexity. The predominant determinant of accuracy was the sample diversity, controlled by the mutation rate $\mu$ in our simulations (cf. SI Fig.~2). For low-diversity samples breakpoints are inferred with accuracy of few basepairs, while for highly diverged isolates most breakpoints are displaced by several hundreds of basepairs (cf. SI Fig.~3). The choice of alignment kernel influences the threshold diversity at which accuracy is lost.

%  TODO: talk aboit the figure
For approximately clonal populations, $\sim 80\%$ of breakpoints were estimated to be within $10$bp away from the ``true" location.
Conversely, for samples with average pairwise diversity above $2\%$, $\sim 80\%$ of breakpoints are estimated only within $100$ bp.
This limitation is primarily driven by lack of sensitivity in the pairwise alignment phase performed using \emph{minimap2} \cite{li2018minimap2}.

\begin{figure}[htb]
    \includegraphics[width=.4\textwidth]{figs/misplaced_fraction_vs_divergence.pdf}
    \caption{{\bf Accuracy against synthetic data}.
        We generate artificial data with varying degree of sequence divergence, and compare the real underlying pangenome graph with the one reconstructed by \textit{PanGraph}, for three different alignment kernels :\textit{minimap2} with \textit{asm10} and \textit{asm20} option, and \textit{mmseqs2}. In each comparison we evaluate the misplacement of breakpoints that we can pair on the two graphs within 1kbp. The plot displays the fraction of breakpoints that have misplacement greater than the standard \textit{PanGraph} precision threshold of 100bp, as a function of average pairwise sequence divergence. Line and shaded area represents mean and standard deviation over 25 repetitions.
        Graph reconstruction with \textit{minimap2} and \textit{asm20} is accurate up to around 5\% average sequence divergence, while for \textit{mmseqs2} accuracy is maintained until $\sim$12\% divergence.
    }
    \label{fig:toy-accuracy}
\end{figure}

@@

\begin{figure*}[htb]
    \includegraphics[width=.9\textwidth]{figs/incr_size.pdf}
    \caption{{\bf Pangenome graph properties vs. dataset size}.
        We build pangenome graphs with increasing number of isolates from the E.~Coli dataset and measure the scaling of different properties of the graphs. Lines and shaded areas represent mean and standard deviation over 10 different repetitions on random subsets of isolates, except for the final point indicating the full graph (307 isolates).
        The number and size of pancontigs scales sub-linearly with the number of isolates, with most of the pangenome being included in ~10\% of the pancontigs, and core pancontigs having higher-than-average size. \textit{PanGraph} is able to compress hundreds of genomes in a pangenome graph having only a fraction of the size.
        \textbf{A}: n. of pancontigs in the graph. We count the total number of pacontigs (blue), the number of core pancontigs (orange) and the minimum number of pancontigs that contain more than 50\% of the pangenome (L50, black).
        \textbf{B}: average size of pancontigs (blue), of only core pancontings (orange), and size of the smallest panconting in the minimal set that spans 50\% of the pangenome (N50, black).
        \textbf{C}: cumulative size of all genomes in the pangenome graph (gray), total pangenome size (blue) and size of core pangenome (orange).
    }
    \label{fig:panx-size}
\end{figure*}

@@

The computational time and maximum memory required to build each alignment is summarized in Table \ref{table:panx-performance}.

\begin{table}[hb]
    \caption{{\bf Real-world Performance}.
        Elapsed time and maximum RAM usage needed to compute the pangraph for the following collections of RefSeq genomes.
        All runs were performed on an AMD Ryzen 7 2700X utilizing 8 cores.
    }
    \begin{tabular}{l r r r }
        \hline\hline
        Species         & Sample size & Wall time (min) & Max memory (GB) \\
        \hline
        E.~coli         & 307         & 426             & 11.7            \\
        H.~pylori       & 85          & 102             & 1.2             \\
        K.~pneumoniae   & 109         & 132             & 2.1             \\
        M.~tuberculosis & 51          & 91              & 0.4             \\
        P.~marinus      & 10          & 12              & 0.1             \\
        \hline
    \end{tabular}
    \label{table:panx-performance}
\end{table}

@@

>>>>

We quantified the accuracy of PanGraph algorithm by computing the displacement of the inferred breakpoints relative to their known locus stored from the evolutionary simulation (cf. SI Section IIB for details). We evaluate this displacement by generating datasets with different rates of mutation $\mu$ and HGT $h$. For each $(h,\mu)$ pair we perform 25 different simulations, and build pangenome graphs using three different options for the alignment kernel, namely \textit{minimap2} \cite{li2018minimap2} with \textit{asm10} and \textit{asm20} option and \textit{mmseqs2} \cite{steinegger2017mmseqs2}.
Critically, we found that the accuracy was \emph{independent} of the rate of horizontal gene transfer $h$ and thus underlying graph complexity. The predominant determinant of accuracy was the sample diversity, controlled by the mutation rate $\mu$ in our simulations (cf. SI Fig.~2). For low-diversity samples breakpoints are inferred with accuracy of few basepairs, while for highly diverged isolates most breakpoints are displaced by several hundreds of basepairs (cf. SI Fig.~3). The choice of alignment kernel influences the threshold diversity at which accuracy is lost.\\
For each alignment kernel and value of average sequence divergence in the population we evaluated the fraction of breakpoints that have displacement greater than the default precision threshold for Pangraph of 100 bp, displayed in Fig.~\ref{fig:toy-accuracy}. On the sequences generated by our simulations, \textit{minimap2} with option \textit{asm20} show a loss of accuracy at a threshold pairwise divergence of around 5\%, while \textit{mmseqs2} is accurate up to average sequence divergence of around 12\%, at the cost of higher computational time.

\begin{figure}[htb]
    \includegraphics[width=.4\textwidth]{figs/misplaced_fraction_vs_divergence.pdf}
    \caption{{\bf Accuracy against synthetic data}.
        We generate artificial data with varying degree of sequence divergence, and compare the real underlying pangenome graph with the one reconstructed by \textit{PanGraph}, for three different alignment kernels :\textit{minimap2} with \textit{asm10} and \textit{asm20} option, and \textit{mmseqs2}. In each comparison we evaluate the misplacement of breakpoints that we can pair on the two graphs within 1kbp. The plot displays the fraction of breakpoints that have misplacement greater than the standard \textit{PanGraph} precision threshold of 100bp, as a function of average pairwise sequence divergence. Line and shaded area represents mean and standard deviation over 25 repetitions. \textit{mmseqs2} maintains accuracy at higher divergence, at the cost of more computational time.
    }
    \label{fig:toy-accuracy}
\end{figure}

@@

\subsubsection*{Dataset characterization}

\subsubsection*{Benchmark}

@@

\subsubsection*{Scaling with dataset size}

\section{Graph marginalization}

@@

% \subsection{Validation on real data}

% We additionally validated PanGraph on natural populations sampled from RefSeq \cite{o2016reference}.
% Obviously the ``true" breakpoints are not known \emph{a priori} and thus the direct analog of Fig. \ref{fig:toy-accuracy} is not possible.
% However, we postulate that most \emph{bona fide} breakpoints generated from evolutionary events will happen at or near gene boundaries; there is preliminary evidence for this assumption \cite{oliveira2017chromosomal}.
% We compared the empirical distribution of distances to gene boundaries to a null distribution generated from the data, in which breakpoints are sampled uniformly from each genome, constrained to produce equivalent number of partitions as was measured by PanGraph.
% For all tested cases, we find our boundaries cluster closer to genes than would be expected by chance.
% The median distance from a Pangraph boundary to annotated gene ranges between 50-100 bp, depending on the species, while the random breakpoint is, on average, 600 bp off.

% \begin{figure*}[htb]
%     \includegraphics[width=.9\textwidth]{figs/incr_size.pdf}
%     \caption{{\bf Pangenome graph properties vs. dataset size}.
%         We build pangenome graphs with increasing number of isolates from the E.~Coli dataset and measure the scaling of different properties of the graphs. Lines and shaded areas represent mean and standard deviation over 10 different repetitions on random subsets of isolates, except for the final point indicating the full graph (307 isolates).
%         The number and size of pancontigs scales sub-linearly with the number of isolates, with most of the pangenome being included in ~10\% of the pancontigs, and core pancontigs having higher-than-average size. \textit{PanGraph} is able to compress hundreds of genomes in a pangenome graph having only a fraction of the size.
%         \textbf{A}: n. of pancontigs in the graph. We count the total number of pacontigs (blue), the number of core pancontigs (orange) and the minimum number of pancontigs that contain more than 50\% of the pangenome (L50, black).
%         \textbf{B}: average size of pancontigs (blue), of only core pancontings (orange), and size of the smallest panconting in the minimal set that spans 50\% of the pangenome (N50, black).
%         \textbf{C}: cumulative size of all genomes in the pangenome graph (gray), total pangenome size (blue) and size of core pangenome (orange).
%     }
%     \label{fig:panx-size}
% \end{figure*}

% % TODO: talk more extensively about the meaning of the results?
% \begin{figure*}[htb]
%     \includegraphics[width=\textwidth]{figs/panx_benchmark.pdf}
%     \caption{{\bf Benchmark on real data}.
%         We build pangenome graphs from fully-assembled chromosomes from 5 different bacterial species. For each species we build graphs with three different alignment kernel options (\textit{minimap2} with \textit{asm10} or \textit{asm20} option and \textit{mmseqs2}) and two different settings for the pseudo-energy parameters (standard or null values).
%         \textbf{A}: \textit{PanGraph} wall-time when run in parallel on 8 cores.
%         \textbf{B}: fraction of core pangenome in the pangenome graph.
%         \textbf{C}: sequence compression, measured as the ratio between the pangenome graph size and the cumulative size of all the sequences contained in the graph.
%     }
%     \label{fig:panx-benchmark}
% \end{figure*}

% The computational time and maximum memory required to build each alignment is summarized in Table \ref{table:panx-performance}.

% \begin{table}[hb]
%     \caption{{\bf Real-world Performance}.
%         Elapsed time and maximum RAM usage needed to compute the pangraph for the following collections of RefSeq genomes.
%         All runs were performed on an AMD Ryzen 7 2700X utilizing 8 cores.
%     }
%     \begin{tabular}{l r r r }
%         \hline\hline
%         Species         & Sample size & Wall time (min) & Max memory (GB) \\
%         \hline
%         E.~coli         & 307         & 426             & 11.7            \\
%         H.~pylori       & 85          & 102             & 1.2             \\
%         K.~pneumoniae   & 109         & 132             & 2.1             \\
%         M.~tuberculosis & 51          & 91              & 0.4             \\
%         P.~marinus      & 10          & 12              & 0.1             \\
%         \hline
%     \end{tabular}
%     \label{table:panx-performance}
% \end{table}

% \begin{figure*}[htb]
%     \includegraphics[width=\textwidth]{figs/marginalize_test.pdf}
%     \caption{{\bf Test of graph marginalization}.}
%     \label{fig:marginalization}
% \end{figure*}
